# Data: Numerical and graphical exploration {.tabset}

## Summary Stats


A summary of the data columns:
```{r}
summary(thrush_sf)
```

Most of them aren't especially informative, but the `VATH` columns tells us that the mean number of thrushes observed was only 0.118.
```{r}
summary(thrush_sf$VATH)
```
There is a pretty big range of elvations in the sample sites:
```{r}
range(thrush_sf$elevation)
```
Some graphical summaries may be more informative.


## Pairplots

We can look to see if any of the terrain variables are correlated:

```{r, echo = FALSE}
pairs.panels(data.frame(thrush_sf)[, c("elevation", "slope", "aspect")])
```


# Create ggplot objects {.tabset}

## Histograms

### Count per transect
```{r, echo = FALSE}
g_hist_census = 
  ggplot(
    aggregate(VATH ~ TRANSECT, data = thrush_sf, sum), aes(x = VATH)) + 
  gm_hist_1 +
  labs(
    title = "Varied Thrush Census", 
    subtitle = "Counts aggregated by transect") +
  xlab("Census Count")

print(g_hist_census)
```


### Terrain Histograms
```{r, echo = FALSE}

gm_hist_thrush = ggplot(thrush_sf) + gm_thrush_hist_2
g_hist_elev = 
  gm_hist_thrush + aes(x = elev) +
  ggtitle("Sample Point Terrain: Elevation") + 
  xlab("Elevation (m)")

g_hist_slope = 
  gm_hist_thrush +  aes(x = slope) +
  ggtitle("Sample Point Terrain: Slope") + 
  xlab("Slope (m / km)")

g_hist_aspect = 
  gm_hist_thrush + aes(x = aspect) + 
  ggtitle("Sample Point Terrain: Aspect") + 
  xlab("Aspect (radians)")

plot_grid(g_hist_elev, g_hist_slope, g_hist_aspect, ncol = 1)
```

### Conditional terrain histograms

Are the distributions of the terrain variables similar at sites with and without thrushes?

```{r}
gm_hist_thrush_present = ggplot(subset(thrush_sf, VATH == 1)) + gm_thrush_hist_3
gm_hist_thrush_absent = ggplot(subset(thrush_sf, VATH == 0)) + gm_thrush_hist_3

g_hist_elev_present = 
  gm_hist_thrush_present + aes(x = elev) +
  ggtitle("Sample Point Elevation: Present") + 
  xlab("Elevation (m)")

g_hist_elev_absent = 
  gm_hist_thrush_absent + aes(x = elev) +
  ggtitle("Sample Point Elevation: Absent") + 
  xlab("Elevation (m)")

plot_grid(g_hist_elev_present, g_hist_elev_absent, nrow = 1)
```