# Data {.tabset}

## Packges, environment

Packages used in this walkthrough.

<div class="warn"> Note:  I use `here()` in conjunction with the file path to the F&F data on my computer. You'll need to tell R where to look for the data on your machine. </div>


<div class="fold s">
```{r packages, results = "hide", warning = FALSE}
```
</div>



## Elevation data

I'm initially reading the raster with `raster()` rather than the stars.
I'm sure there's a stars-friendly pre-built terrain function, but I haven't found it yet.

You can use `terrain_brick()` from package `raster` to calculate slope and aspect.

Finally, group everythign into a multilayer raster `stars` object.

<div class="fold s">
```{r elevation_data, results = "hide", warning = FALSE}
```
</div>




## Varied Thrush Data

- I prefer opening tabular data files with `data.table`, it 'just works' much more often than `read.csv()`.
- You can make a `sf` object from a `data.frame` or `data.table` with `st_as_sf()`
- I have been unsuccessful at finding a quick way to extract values from a stars raster brick given point coordinates.
    - This code uses `extract()` on the `RasterBrick` version of the data to add terrain info to the thrush data.

<div class="fold s">
```{r varied_thrush_data, results = "hide", warning = FALSE}
```
</div>




## Center data

The authors centered and scaled, i.e. a z-transfomation, for the modeling.  The resulting
centered/scaled variables have mean 0 and variance 1.

```{r center_data, results = 'hide', warning = FALSE}
```


