---
title: "Fletcher and Fortin, Chapter 6"
subtitle: "Varied Thrush Example Walkthrough"
output: 
  # pdf_document:
    # toc: yes
  html_document:
    css: !expr here::here("css", "styles.css")
    number_sections: TRUE
    toc: true
    toc_float: true
---

```{r setup, include = FALSE, echo = FALSE}

require(here)
require(stars)
require(sf)
require(ggplot2)
require(viridis)
require(spData)
require(raster)
require(cowplot)
require(sp)

source(here("/data/environment_vars.R"))

# require(raster)
# require(psych)


# require(spatstat)
# require(lulcc)
```


# Data and helper functions

```{r ggplot graphics objects}
gg_rast = ggplot() + 
  coord_equal() + 
  scale_fill_viridis(
    na.value = "transparent",
    guide = 
      guide_colorbar(
        title.position = "top",
        barwidth = unit(0.6, "npc"))) + 
  scale_x_discrete(expand=c(0,0)) +
  scale_y_discrete(expand=c(0,0)) +
  theme_map() +
  theme(
    legend.position = "bottom", 
    legend.direction = "horizontal",
    axis.title = element_blank())

gg_rast_gs = ggplot() + 
  coord_equal() + 
  scale_fill_viridis_c(
    na.value = "transparent",
    guide = 
      guide_colorbar(
        title.position = "top",
        barwidth = unit(0.6, "npc"))) + 
  scale_x_discrete(expand=c(0,0)) +
  scale_y_discrete(expand=c(0,0)) +
  theme_map() +
  theme(
    legend.position = "bottom", 
    legend.direction = "horizontal",
    axis.title = element_blank())


```

# Elevation/Terrain Data
```{r}
 dem_crs_m = "+proj=aea +lat_1=46 +lat_2=48 +lat_0=44 +lon_0=-109.5 +x_0=600000 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
  
  dem = raster(file.path(book_data, "elev"))
  proj4string(dem) = dem_crs_m
  
  dem_m = dem
  dem_m[] = dem[] * 1e3
  
  dat_brick = brick(dem_m, terrain(dem_m, opt = c("slope", "aspect"), unit = "radians"))
  names(dat_brick)[1] = "elevation"
  dat_stars = st_as_stars(dat_brick)
  
  rm(dem, dem_m) 
```

# Thrush Data

```{r}
thrush_pts <- read.csv(file.path(book_data, "vath_2004.csv"), header = TRUE)
  
  thrush_pts = 
    within(thrush_pts,
           {
             census = factor(VATH)
             levels(census) = c("absent", "present")
           }
    )
  head(thrush_pts)
  coordinates(thrush_pts) = ~ EASTING + NORTHING
  proj4string(thrush_pts) = dem_crs_m
  
  thrush_sf = st_as_sf(thrush_pts)
```






## Elevation/terrain data
```{r data import}

dem_crs_m = "+proj=aea +lat_1=46 +lat_2=48 +lat_0=44 +lon_0=-109.5 +x_0=600000 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"

dem = raster(file.path(book_data, "elev"))
proj4string(dem) = dem_crs_m
```

The book's supplied raster file has elevation in km, but it's easier to work with meters.
```{r}

dem_m = dem
dem_m[] = dem[] * 1e3

dat_terrain = brick(dem_m, terrain(dem_m, opt = c("slope", "aspect"), unit = "radians"))
names(dat_terrain)[1] = "elevation"
dat_rast = st_as_stars(dat_terrain)

dat_rast[, , 1]

g_elev = gg_rast + geom_stars(data = dat_rast[, , , 1]) + labs(fill = "Elevation")
g_slop = gg_rast + geom_stars(data = dat_rast[, , , 2]) + labs(fill = "Slope")
g_aspt = gg_rast + geom_stars(data = dat_rast[, , , 3]) + labs(fill = "Aspect")

```

## Plot terrain
```{r}

if(FALSE)
{
  
  map_width = 6.5
  map_height = 8
  map_res = 250
  png(file.path(fig_dir, "thrush_elevation.png"), 
      width = map_width, height = map_height, units = "in", res = map_res)
  print(g_elev)
  dev.off()
  
  png(file.path(fig_dir, "thrush_aspect.png"), 
      width = map_width, height = map_height, units = "in", res = map_res)
  print(g_aspt)
  dev.off()
  
  png(file.path(fig_dir, "thrush_slope.png"),
      width = map_width, height = map_height, units = "in", res = map_res)
  print(g_slop)
  dev.off()
  
  png(file.path(fig_dir, "thrush_terrain.png"), 
      width = 3 * map_width, height = map_height, units = "in", res = map_res)
  print(plot_grid(g_elev, g_slop, g_aspt, nrow = 1))
  dev.off()
}

```


## Thrush data


```{r}
thrush_pts <- read.csv(file.path(book_data, "vath_2004.csv"), header = TRUE)

thrush_pts = 
  within(thrush_pts,
         {
           `census` = factor(VATH)
           levels(census) = c("absent", "present")
         }
  )
head(thrush_pts)
plot(x = EASTING, y = NORTHING, data = thrush_pts)

with(thrush_pts,
     
     plot(NORTHING, EASTING)
)

locator(2)

coordinates(thrush_pts) = ~ EASTING + NORTHING
proj4string(thrush_pts) = dem_crs_m
plot(thrush_pts)


thrush_sf = st_as_sf(thrush_pts)

```





# Overview

F + F use example data from the **Northern Region Landbird Monitoring Program**.
> Hutto RL, Young JS (2002) Regional landbird monitoring: perspectives from the Northern Rocky Mountains. Wildl Soc Bull 30( 3): 738– 750


## The data

Surveys for the varied thrush: *Ixoreus naevius*

- Sampling scheme:
- ~3km transects
- point surveys - 100m radius
- 10 pts/transect

## Authors' correlogram function

Note that there is a mistake in the definition of the `moran.results` data frame.

They define two columns called `null.lower`, however one of them should have been `null.upper`

```{r}
icorrelogram <- function(locations, z, binsize, maxdist)
{
  # package 'spdep' needed for dnearneigh() function.
  # Check whether it is installed:
  stopifnot(require(spdep))
  
  distbin <- seq(0, maxdist, by = binsize)
  Nbin <- length(distbin)-1
  
  moran.results <- data.frame(
    "dist" =  rep(NA, Nbin), 
    "Morans.i" = NA, 
    "null.lower" = NA, 
    "null.upper" = NA)
  
  for (i in 1:Nbin){
    d.start <- distbin[i] 
    d.end <- distbin[i+1]
    
    neigh <- spdep::dnearneigh(
      x = locations, 
      d1 = d.start, 
      d.end, 
      longlat = F)
    
    wts <- nb2listw(
      neighbours = neigh, 
      style = 'B', 
      zero.policy = T)
    
    #note alternative is for P-value, so only 'significant if positive autocorrelation
    mor.i <- moran.mc(
      x = z, listw = wts, 
      nsim = 200, 
      alternative = "greater", 
      zero.policy = T) 
    
    moran.results[i, "dist"] <- (d.end+d.start)/2 
    
    #observed moran's i
    moran.results[i, "Morans.i"] <- mor.i$statistic 								                
    
    moran.results[i, "null.lower"] <- 
      quantile(
        mor.i$res, 
        probs = 0.025, 
        na.rm = T)#95% null envelope	
    
    moran.results[i, "null.upper"] <- 
      quantile(
        mor.i$res, 
        probs = 0.975, 
        na.rm = T) #95% null envelope
  }
  return(moran.results)
}

# plotting function
#
plot_ff_correlogram = function(cgram, title, subttl = "", thm = theme_light(), ci_lty = 2)
{
  require(ggplot2)
  return(
    ggplot(cgram, aes(x = dist)) +
      geom_line(aes(y = Morans.i)) +
      geom_line(aes(y = null.upper), lty = ci_lty) +
      geom_line(aes(y = null.lower), lty = ci_lty) +
      ylab("Moran's I") + xlab("distance") +
      ggtitle(title, subttl) +
      geom_hline(yintercept = 0, lty = 3)
  )
}

```



# Thrush Predictor Data
```{r}



```





# Data import

## Elevation data

Northern Idaho, western Montana.
```{r}
# elev = raster(file.path(book_data, "elev.gri"))
# elev = raster(file.path(book_data, "elev.grd"))
elev = raster(file.path(book_data, "elev"))
```


Compute slope and aspect using `raster` package.

<div class = "red"> NOTE: The second line of code in the example provided in the text fails:

```{r, error = TRUE}
elev.terr <- terrain( elev, opt = c(" slope", "aspect"), unit = "radians")
```

They neglect to mention that the raster they provide does not include projection information:
```{r}
proj4string(elev)=CRS(" +proj=aea +lat_1=46 +lat_2=48 +lat_0=44 +lon_0=-109.5 +x_0=600000 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")
```
</div>


Build terrain layers using `terrain()` in the `raster` package

Note: the book code specifies the unit as radians, the supplemental material code does not:
```{r}
# Supplemental code version
# elev.terr <- terrain(elev, opt = c("slope", "aspect"))

# Book version
elev.terr <- terrain(elev, opt = c("slope", "aspect"), unit = "radians")
```

Create a raster stack to hold all the layers:
```{r}
layers <- stack(elev, elev.terr)
layers <- brick(elev, elev.terr)
names(layers) <- c("elev", "slope", "aspect")
class(layers)
```


You can use `subset()` to select a particular layer i n a `RasterStack` or `RasterBrick` object.
```{r}
par(mfrow = c(1, 3))
plot(subset(layers, 1))
plot(subset(layers, 2))
plot(subset(layers, 3))
```



## Varied Thrush Data

Import and plot:
```{r}
#Point data with presence/absence
point.data <- read.csv(file.path(book_data, "vath_2004.csv"), header = TRUE)

dev.off()
plot(subset(layers, 1))

#plots the presence points
points(point.data[point.data$VATH == 0, c("EASTING", "NORTHING")], col = "black", pch = 21, cex = 0.8)

#plots the absence points
points(point.data[point.data$VATH == 1, c("EASTING", "NORTHING")], col = "red", pch = 16, cex = 0.7)
```

Extract data data at sampling points
```{r}



coords <- cbind(point.data$EASTING, point.data$NORTHING)
land.cov <- extract(x = layers, y = coords)
point.data <- cbind(point.data, land.cov)
```


# Aspatial logistic regression

## Data set-up

### Check for collinearity (correlations among predictors)
```{r}

psych::pairs.padata.frame(thrush_sf)


cor(point.data[, c("elev", "slope", "aspect")], method = "pearson")
# pairs(point.data[, c("elev", "slope", "aspect")])
# the psych package implementatin looks much nicer
psych::pairs.panels(point.data[, c("elev", "slope", "aspect")])
```


### Center and scale predictors for modeling
```{r}
point.data$elevs <- scale(point.data$elev, center = T, scale = T)
point.data$slopes <- scale(point.data$slope, center = T, scale = T)
point.data$aspects <- scale(point.data$aspect, center = T, scale = T)
```



## Model 1: elevation 
```{r}
VATH.elev <- glm(
  VATH ~ elev, 
  family = "binomial", 
  data = point.data)
summary(VATH.elev)
```



## Model 2: additive - all factors
```{r}
VATH.all <- glm(
  VATH ~ elevs + slopes + aspects, 
  family = "binomial", 
  data = point.data)
summary(VATH.all)
```




## Model 3: polynomial elevation
```{r}
VATH.elev2 <- glm(
  VATH ~ elev + I(elev^2), 
  family = "binomial", 
  data = point.data)
summary(VATH.elev2)
```




## Model Selection

The polynomial elevation model has lowest AIC
```{r}
AIC(VATH.elev, VATH.all, VATH.elev2)
```


Extract coefficients and their SEs from the best model
```{r}
summary(VATH.elev2)$coef
glm.summary <- c(
  summary(VATH.elev2)$coef[2, 1], 
  summary(VATH.elev2)$coef[2, 2], 
  summary(VATH.elev2)$coef[3, 1], 
  summary(VATH.elev2)$coef[3, 2])
#inspect
glm.summary
```
<br><br>


# Plot (aspatial) model

## Use model to plot predicted Pr(abundance)

First create a new (aspatial) data set to use with `predict()`
```{r}
Elev <-  seq(
  min(point.data$elev), 
  max(point.data$elev), 
  length = 15)
newdata <- data.frame(elev = Elev)
```

Use model to predict values for different elevations:
```{r}
# type = response for predicted probabilities
glm.pred <- predict(
  VATH.elev2, 
  newdata = newdata, 
  type =  "link", 
  se = T) 
```

Logistic regression fits are given in quantiles of the **logistic distribution**, but these numbers aren't very intuitive to interpret.  We can back-transform them onto a probability scale:
```{r}
glm.newdata <-
  data.frame(
    newdata, 
    pred = plogis(glm.pred$fit), 
    lcl = plogis(glm.pred$fit + 1.96*glm.pred$se.fit), 
    ucl = plogis(glm.pred$fit - 1.96*glm.pred$se.fit))
```

## Plot model coeff. with confidence intervals
```{r}
with(
  glm.newdata,
  matplot(
    x = elev, y = cbind(pred, lcl, ucl), 
    type = "l", col = 1, lty = c(1, 2, 2),
    xlab = "elevation", ylab = "coeff"),
  ylim = c(0, 0.3)
)
```


## Map of model predictions

The `raster` package includes an implementation of `predict` that we can use to create a raster layer of model predictions:
```{r}
glm.raster <- predict(
  model = VATH.elev2,
  object = layers,
  type = "response")
plot(glm.raster, xlab = "Longitude", ylab = "Latitude")
```



# Spatial dependence of model output


## Correlogram of response

Create a correlogram using the authors' function:
```{r}
VATH.cor <- icorrelogram(
  locations = coords, 
  z = point.data$VATH, 
  binsize = 1000, 
  maxdist = 15000)
VATH.cor
```


<div class = "red">NOTE: the authors had an error in the `lines()` call below.  They misspecified the upper and lower null envelope column names in the original.</div>

```{r}
round(head(VATH.cor, 3), 2)
plot_ff_correlogram(VATH.cor, "Correlogram of Response", "elevation-polynomial model")
```


## Correlogram of model residuals

Build a correlogram with the authors' function:
```{r}
VATH.elev2.res <- residuals(VATH.elev2, type = "deviance")
corr.res <- icorrelogram(
  locations = coords, 
  z = VATH.elev2.res, 
  binsize = 1000, 
  maxdist = 15000)
```

<div class = "red"> NOTE: same comment as above about error in column names.</div>
```{r}

plot_ff_correlogram(corr.res, "Correlogram of Residuals", "elevation-polynomial model")
```



contrast results from raw to residuals with mean
correlogram on residuals of mean model (intercept model)
```{r}
VATH.int <- glm(VATH ~ 1, family = "binomial", data = point.data)
VATH.int.res <- residuals(VATH.int, type = "deviance")
corr.int.res <- icorrelogram(locations = coords, z = VATH.int.res, binsize = 1000, maxdist = 15000)

plot_ff_correlogram(corr.int.res, "Correlogram of Residuals", "intercept-only model")
```

correlation
```{r}
cor(VATH.cor$Morans.i, corr.int.res$Morans.i)
```


# Subset data to account for autocorrelation ----
```{r}
#randomly shuffle data by transect and create a shuffled rank vector
rand.vector <- with(
  point.data, 
  ave(
    POINT, 
    as.factor(TRANSECT), 
    FUN = function(x) {sample(length(x))}))

#pick one random point on transect and remove rest
point.datasub <- point.data[rand.vector  <=  1, ]
head(point.datasub, 3)

```

coordinates subset

```{r}
coords.sub <- cbind(point.datasub$NORTHING, point.datasub$NORTHING)
head(coords.sub, 3)
#model
VATH.sub <- glm(VATH~elev + I(elev^2), family = "binomial", data = point.datasub)
summary(VATH.sub)
```

extract coefficients
```{r}
glmsub.summary <- 
  c(
    summary(VATH.sub)$coef[2, 1], 
    summary(VATH.sub)$coef[2, 2], 
    summary(VATH.sub)$coef[3, 1], 
    summary(VATH.sub)$coef[3, 2])
```

residuals
#correlogram on residuals
```{r}
VATH.sub.res <- residuals(VATH.sub, type = "deviance")
corr.sub.res <- icorrelogram(locations = coords.sub, z = VATH.sub.res, binsize = 2000, maxdist = 15000)
plot_ff_correlogram(corr.sub.res, "Correlogram of Residuals", "spaced-sample model")
```
