# Data: Import and setup {.tabset}

## Packges, environment

Packages used in this walkthrough.

<div class="warn"> Note:  I use `here()` in conjunction with the file path to the F&F data on my computer. You'll need to tell R where to look for the data on your machine. </div>
<br>

<div class="fold s">

```{r packages, eval = FALSE, class.source='klippy'}
```
</div>




## Elevation data

I'm initially reading the raster with `raster()` rather than the stars.  
I'm sure there's a stars-friendly pre-built terrain function, but I haven't found it yet.

You can use `terrain_brick()` from package `raster` to calculate slope and aspect.

Finally, group everythign into a multilayer raster `stars` object.


<div class="fold s">

```{r elevation data, eval = FALSE, class.source='klippy'}
```
</div>




## Varied Thrush Data

- I prefer opening tabular data files with `data.table`, it 'just works' much more often than `read.csv()`.
- You can make a `sf` object from a `data.frame` or `data.table` with `st_as_sf()`
- I have been unsuccessful at finding a quick way to extract values from a stars raster brick given point coordinates.
    - This code uses `extract()` on the `RasterBrick` version of the data to add terrain info to the thrush data.

<div class="fold s">
```{r varied thrush data, eval = FALSE, class.source='klippy'}
```
</div>



## Center data

<div class="fold s">
```{r center data, eval = FALSE, class.source='klippy'}
```
</div>




## Correlogram functions

This is my version of the authors' function for calculating a correlogram.  

<div class="warn">Note that there is a mistake in the definition of the `moran.results` data frame.

They define two columns called `null.lower`, however one of them should have been `null.upper`</div>

<div class="fold s">
```{r correlogram function, class.source='klippy'}
```
</div>




## Plot correlogram
There's also a convenience function for plotting the resulting correlogram.
<div class="fold s">
```{r plotting function, eval = FALSE, class.source='klippy'}
```
</div>





## Graphical paramaters

- Create template ggplot objects
- Set image file output params

<div class="fold s">
```{r plotting params, eval = FALSE, class.source='klippy'}
```
</div>



## Plotting {.tabset}


<!-- ### Pair plot -->
<!-- ```{r covariate pair plot, eval = FALSE, class.source='klippy'} -->
<!-- ``` -->

### Census histograms
```{r census histograms, eval = FALSE, class.source='klippy'}
```

### Terrain histograms
```{r terrain histograms, eval = FALSE, class.source='klippy'}
```

### Thrush map
```{r thrush map, eval = FALSE, class.source='klippy'}
```



