# Data: Numerical and graphical exploration {.tabset}

## Summary Stats

A summary of the data columns:
<div class="fold o">
```{r}
summary(thrush_sf)
```
</div>


That's a lot of info, most of which isn't especially helpful for our purposes.

The `count` column tells us that the mean number of thrushes observed was only 0.118.
```{r}
summary(thrush_sf$count)
```


There is a pretty big range of elvations in the sample sites:
```{r}
range(thrush_sf$elevation)
```

Some graphical summaries will be informative.



## Histograms

### Count per transect

A histogram of the number of thrushes observed per **transect**.  Remember that there were multiple **sample points** along each transect.
```{r, echo = FALSE}
print(g_hist_census)
```


### Terrain Histograms

How are the terrain covariates distributed among the sample point locations?

```{r, echo = FALSE}
plot_grid(g_hist_elev, g_hist_slope, g_hist_aspect, ncol = 1)
```



### Conditional terrain histograms

Are the distributions of the terrain variables similar at sampling points where thrushes were **present** and **absent**?

```{r}
plot_grid(g_hist_elev_present, g_hist_elev_absent, nrow = 1)
```




## Pairplots

We can look to see if any of the terrain variables are correlated:

```{r covariate_pair_plot}
```

