
# Data {.tabset}


## Packges, environment

Packages used in this walkthrough.

<div class="warn"> Note:  I use `here()` in conjunction with the file path to the F&F data on my computer. You'll need to tell R where to look for the data on your machine. </div>


<div class="fold s">
```{r packages, results = "hide", warning = FALSE}
```
</div>



## Elevation data

Read in the elevation using `raster()`

```{r elevation_data, results = "hide", warning = FALSE}
```


## Build terrain covariates

You can use `terrain_brick()` from package `raster` to calculate slope and aspect.

```{r build_terrain}
```




## Varied Thrush Data {.tabset}

First read the data into a data frame, then make a spatial object:

```{r read_thrush_data_frame}
```


### `sf` version

- You can make a `sf` object from a `Spatial`, `data.frame`, or `data.table` object with `st_as_sf()`
```{r read_thrush_data_frame, results = 'hide'}
```
```{r thrush_data_frame_to_sf }
```



### `SpatialPointsDataFrame` version
- Raster works better with `Spatial` objects than `sf` objects, so I'll proceed with the SPDF version.

```{r read_thrush_data_frame}
```
```{r thrush_spdf}
```


### Extract terrain

- I have been unsuccessful at finding a quick way to extract values from a `stars` object given point coordinates.
    - This code uses `extract()` on the `RasterBrick` version of the data to add terrain info to the thrush data.

```{r extract_terrain_data}
```

- The rest of the code works better with an `sf` object:
```{r convert_to_sf, results = "hide"}
```



### Center data

The authors centered and scaled, i.e. a z-transfomation, for the modeling.  The resulting
centered/scaled variables have mean 0 and variance 1.

```{r center_data, results = 'hide', warning = FALSE}
```


