# Aspatial model building {.tabset}

Let's try three aspatial **logistic regression models**:

We'll use three different sets of predictors.

## Elevation
<div class="fold o">
```{r model_1_center, class.source='klippy'}
```
</div>



## Elevation, slope, aspect
<div class="fold o">
```{r model_2_center, class.source='klippy'}
```
</div>




## Quadratic elevation
<div class="fold o">
```{r model_3_center, class.source='klippy'}
```
</div>




# Aspatial Model Diagnostics {.tabset}

In this example, I'm including only some **very** basice model diagnostics: residual plots.  I'm also going to proceed as if the diagnostics were totally fine.  In a real analysis, you'd want to perform a more thorough set of model diagnostics.


Here's a very nice article about plotting residuals with ggplot: 
<a href="https://drsimonj.svbtle.com/visualising-residuals">Visualising Residuals by Simon Jackson</a>

The following are simple base R plots:

## Elevation Model

Elevation vs residuals and predicted vs. residuals plots for model 1: 

<div class="fold s">
```{r model_1_diagnostics}
```
</div>


## Quadratic Elevation Model

We'll only plot the predicted vs. residuals plot here.  Since this model has mutliple predictors plotting predictors vs. residuals is more complicated.

<div class="fold s">
```{r model_3_diagnostics}
```
</div>



# Model Selection

We'll use a very simple model selection process: the Akaike Information Criterion (AIC).

The AIC attempts to balance model complexity with model fit.  A low AIC value indicates a 'better' model.

```{r model_selection}
```

The quadratic elevation model seems to be the best because its AIC is the lowest, by around 17.

<div class="warn"> NOTE: AIC and model selection are complex topics, this is a very simplified selection procedure.  In real life you would perform a more thoughtful and thorough model selection procedure. </div>
