---
title: "Chapter 05 part 1"
author: "Michael France Nelson"
date: "March, 2020"
output: 
  # pdf_document:
  #   toc: yes
  html_document:
    css: !expr here::here("css", "styles.css")
  # css: ../../css/styles.css
number_sections: TRUE
toc: true
toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(here)
require(geoR)
require(gstat)
require(pgirmess)
require(sp)
# require(ncf)

source(here("/data/environment_vars.R"))



# require(raster)
# require(rgdal)
# require(rgeos)
# require(spatstat)
# require(ggplot2)

# knitr::opts_chunk$set(root.dir = here(), error = TRUE)
# knitr::opts_knit$set(echo = TRUE, error = TRUE, root.dir = here())


```


# Correlograms: 5.3.3

<div class="red">
NOTE: The first line of code in the book reads the data into a `data.frame` called `matrix`.  DO NOT do this!  `matrix` is the name of a function.  R allows you to overwrite a function name in this way, but it is **extremely** bad practice.
</div>


## Import and plot vegetation matrix data

It might be nicer if the data were already in a raster format.




```{r read cactus matrix}

file.exists(paste0(book_data, "cactus_matrix.csv"))
cactus_matrix = read.csv(paste0(book_data, "cactus_matrix.csv"))
str(cactus_matrix)

```


Here's a way to make it a spatial points object:

```{r cactus matrix data}
matrix_spdf = cactus_matrix
coordinates(matrix_spdf) = ~ x + y
```



Nice use of the `cut()` function:

```{r plot cactus matrix}
n_breaks = 17
cex_mat = 5
dev.off()
par(mar = c(0, 0, 0, 0))
plot(
  cactus_matrix$x, cactus_matrix$y, 
  pch = 22, col = 0, cex = cex_mat, asp = 1,
  bg = gray.colors(n_breaks)[cut(cactus_matrix[, 3], breaks = n_breaks)], 
  ann = F, axes = F)
```


Using the spatial points object:

```{r plot spdf}
plot(
  matrix_spdf, 
  pch = 22, col = 0, cex = cex_mat, asp = 1,
  bg = gray.colors(n_breaks)[cut(cactus_matrix[, 3], breaks = n_breaks)], 
  ann = F, axes = F)

```


## Build a distance matrix

We know that the measurements are points on a rectangular grid, spaced 2m apart.

The distances in the distance matrix will be cell-wise distances:

```{r distance matrices}

matrix_dist = dist(cbind(cactus_matrix[, c("x", "y")]))
maxdist = (2 / 3) * max(matrix_dist)

matrix_dist_sp = spDists(matrix_spdf)
class(matrix_dist_sp)
```


## Correlogram using package `pgirmess`



```{r pgirmess correlogram}
correlog_pgirmess = pgirmess::correlog(
  coordinates(matrix_spdf), 
  matrix_spdf$Height, 
  method = "Moran", 
  nbclass = 14, 
  alternative = "two.sided")

class(correlog_pgirmess)
summary(correlog_pgirmess)
head(correlog_pgirmess)
dev.off()
plot(correlog_pgirmess)
abline(h = 0, lty = 2)

```


## Correlogram with `ncf`


```{r ncf correlogram}

correlog_ncf = ncf::correlog(
  cactus_matrix$x,
  cactus_matrix$y,
  cactus_matrix$Height,
  increment = 5, resamp = 99)


correlog_spline = ncf::spline.correlog(
  cactus_matrix$x,
  cactus_matrix$y,
  cactus_matrix$Height,
  xmax = maxdist,
  resamp = 100,
  type = "boot")
  

```

```{r plot ncf correlograms}

plot(correlog_ncf)
plot(correlog_spline)

```

```{r}
str(correlog_pgirmess)
str(correlog_ncf$correlation, 0)
str(correlog_spline$real$predicted, 1)

```



