---
title: "R setup"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
require(raster)
book_data = "C:/Users/michaelnelso/Documents/courses/synced/Box Sync/Fletcher_Fortin_data/Fletcher_Fortin-2018-Supporting_Files/data/"
dir.exists(book_data)
```


# 

# Setting up a data folder
Note: You have to download the data for the Fletcher and Fortin textbook.
You can find the file here:

https://ufdc.ufl.edu/IR00010770/00001

Unzip the file into a location that you'll remember.

Make a string variable called *data_folder* that contains the path to your data folder.

You'll need to keep track of this location for all of the labs and exercises that use any of the text's supplemental data.

On my work computer, I saved the data in the following location:

```{r}
book_data = "C:/Users/michaelnelso/Documents/courses/synced/Box Sync/Fletcher_Fortin_data/Fletcher_Fortin-2018-Supporting_Files/data/"
```


Run the following as a quick test to make sure you have stored your data path correctly in R:
```{r}
dir.exists(book_data)
```

If you get *TRUE* you are ready to proceed.  Otherwise re-check the contents of your *book_data* variable.  Note that the file system text conventions may be different from the example I provided if you are using Mac OS or Linux.

# Loading data from Fletcher and Foprtin

To load a dataset from this folder, you'll need to use the paste0() function.  

The syntax for loading data will differ slightly from the code used in the text.

For example, the text provides the following code to load the National Land Cover Database data:
```{r}
nlcd <- raster("nlcd2011SE")
```

Instead, we'll use *paste0()* along with our variable pointing to our data path:
```{r}
file.exists(paste0(book_data, "nlcd2011SE"))
nlcd = raster(paste0(book_data, "nlcd2011SE"))
print(nlcd)



```

