---
title: "Fletcher and Fortin, Chapter 4: Spatial Dispersion and Point Data"
subtitle: "Walkthrough by Michael France Nelson, part 2 - marked point patterns"
output: 
  # pdf_document:
    # toc: yes
  html_document:
    css: ../css/styles.css
    number_sections: TRUE
    toc: true
    toc_float: true
---




```{r setup, include = FALSE, echo = FALSE}
if (! "here" %in% data.frame(installed.packages())$Package) install.packages("here")

# We don't want maptools for this material:
# if (! "maptools" %in% data.frame(installed.packages())$Package) install.packages("maptools")

# require(raster)
# require(rgdal)
# require(rgeos)
# require(ggplot2)

require(sp)
require(spatstat)
require(here)

knitr::opts_chunk$set(root.dir = here(), error = TRUE)
knitr::opts_knit$set(echo = TRUE, error = TRUE, root.dir = here())

source(here("/data/environment_vars.R"))
```


# Overview

I prefer to avoid storing intermediate variables in the R gobal environment if I don't have to.

I like to use them when I'm learning new techniques, writing functions, or giving demos.  

I think final code is more readable and efficient with as few intermediate variables as possible.  
It also helps with debugging and avoiding unintended global variables producing unintended effects debugging: think of the extra variable `i` in the book's `BufferCover()` function from Chapter 2.


I'm going to set the random number generator seed for reproducibalilty:
```{r}
set.seed(1234554321)
```



# Package 'here'

I recently discovered the package 'here'.  It is fantastic.

This code will conditionally install packages if they are not present:

```{r, eval = FALSE}
if (! "here" %in% data.frame(installed.packages())$Package) install.packages("here")
```




# Data import 


## Using `coordinates()`

You can use `coordinates() =` as a shortcut to build a spatial points object from a dataframe.

Note:  this is a R function that *does* use side effects!  R conventions aren't totally consistent on side effects, but it's usually safe to assume that functions do't use them.

`coordinates` can accept a vector of coordinate column names, or you can use the formulat notation.

```{r prepare data, error = TRUE}
cactus = read.csv(paste0(book_data, "/cactus.csv"), header = T)
coordinates(cactus) = ~ East + North

cactus = read.csv(paste0(book_data, "/cactus.csv"), header = T)
coordinates(cactus) = c("East", "North")

```


## `owin()`

We don't need to bother with the import of the cactus boundaries.

Since we've already imported the cactus data, we can just use its bounding box.

The code is a little bit awkward, but it avoids cluttering up the environment with intermediate variables.

```{r}
owin(range(cactus$North), range(cactus$East))
```


## `ppp()` from `SpatialPoints*`

Package `maptools` has tools to build `ppp` objects directly from `SpatialPoints*` objects.

We can avoid creating an `owin` object using this method.

```{r}
require(maptools)
ppp_cactus = as.ppp(cactus)
```


# Marked Point Patterns - section 4.3.5



## Cactus data marks

*Chelinidea vittiger* is a herbivore on *O humifusa*, may be a biocontrol agent.
`ppp` marks can be vectors (for a single mark) or dataframes (for multiple marks).

We want a factor mark to indicate whether or not *C. vittiger* are present.



## Using `within()` and `with()`)

The function `within()` lets you temporarily work with the columns of a `data.frame` as if they were in the global environment.  It's a little like using functins within the indices of a `data.table`.  It returns a `data.frame` with any additions you made.

`with()` is similar, but it doesn't return the entire `data.frame`

These functions help avoid using temporary variables and can make code more readable by cutting down on subsetting notation.

These functions can do more sophisticated tasks too.



### `within()` returns a `data.frame`

```{r}
head(within(
  marks(ppp_cactus),
  {
    ch_pr_ab = factor(chelinidea > 0)
  }
))

marks(ppp_cactus) = 
  within(
    marks(ppp_cactus),
    {
      ch_pr_ab = factor(chelinidea > 0)
    }
  )

head(marks(ppp_cactus))

```


### `with()` returns a vector

```{r}
ppp_cactus = as.ppp(cactus)
head(marks(ppp_cactus))

head(with(
  marks(ppp_cactus),
  {
    factor(chelinidea > 0)
  }
))

marks(ppp_cactus)$ch_pr_ab = with(
  marks(ppp_cactus),
  {
    factor(chelinidea > 0)
  }
)

head(marks(ppp_cactus))

```


### Chelinidea presence/absence

I like the `within()` method better:

```{r}
ppp_cactus = as.ppp(cactus)

marks(ppp_cactus) = 
  within(
    marks(ppp_cactus),
    {
      ch_pr_ab = factor(chelinidea > 0)
    }
  )
```


## PPP Summary Statistics

The text shows `split()` to print summary statistics, but this
doesn't tell us much other than point counts.  You can use 
`summary(split())` on a `ppp` with multiple marks for more info:

```{r}
split(ppp_cactus)
summary(split(ppp_cactus))

```



# Univariate L-Functions

## Univarite *Opuntia*

We can re-do the univariate point pattern analyses using the marked cactus ppp.

`unmark()` strips the marks from a `ppp`.



```{r}

l_env_cactus = envelope(
  unmark(ppp_cactus), 
  fun = Lest,
  nsim = 99,
  rank = 1,
  correction = "iso",
  global = FALSE,
  verbose = FALSE)
```

Plot to verify that these look the same:

```{r}
plot(l_env_cactus)
plot(l_env_cactus, . - r ~ r)
```

## Univariate insect

```{r}

l_env_cheli = envelope(
  unmark(subset(ppp_cactus, ch_pr_ab == "TRUE")),
  fun = Lest,
  nsim = 99,
  rank = 1,
  correction = "isotropic",
  global = FALSE,
  verbose = FALSE)

plot(l_env_cheli, . - r ~ r, legend = FALSE)

```


# Bivariate L function

With random-labeling:

I haven't found a function that can strip selected marks from a `ppp` but it probably exists.

For now, I'll make an unmarked `ppp` object to work with.

```{r}
ppp_cheli = unmark(ppp_cactus)
marks(ppp_cheli) = marks(ppp_cactus)$ch_pr_ab

l_cross_1 =
  envelope(
    ppp_cheli, Lcross,
    nsim = 99, rank = 1, 
    i = "FALSE",
    global = FALSE,
    simulate = expression(rlabel(ppp_cheli)))

l_cross_2 =
  envelope(
    ppp_cheli, Lcross,
    nsim = 99, rank = 1, 
    i = "TRUE",
    global = FALSE,
    simulate = expression(rlabel(ppp_cheli)))

plot(l_cross_1, . - r ~ r)
plot(l_cross_2, . - r ~ r)

```

# Mark Correlation Function


```{r}

ppp_cactus
ppp_cactus_area = unmark(ppp_cactus)
marks(ppp_cactus_area) = marks(ppp_cactus)$Area

plot(markcorr(ppp_cactus_area))

mcf_env = envelope(
  ppp_cactus_area,
  markcorr,
  nsim = 99,
  correction = "iso",
  global = FALSE)

plot(mcf_env, shade = c("hi", "lo"), legend = FALSE)


```

# Inhomogeneous Process Models



```{r}
ppm_intercept = ppm(unmark(ppp_cactus), ~1)
ppm_linear    = ppm(unmark(ppp_cactus), ~ x + y)
ppm_quad      = ppm(unmark(ppp_cactus), ~ polynom(x, y, 2))

ppm_intercept
ppm_linear
ppm_quad

plot(ppm_intercept)

str(ppm_intercept, 1)

plot(ppm_linear)




```


# Simulating Point Patterns




