---
title: "Fletcher and Fortin, Chapter 4: Spatial Dispersion and Point Data"
subtitle: "Walkthrough by Michael France Nelson"
output: 
  # pdf_document:
    # toc: yes
  html_document:
    css: ../css/styles.css
    number_sections: TRUE
    toc: true
    toc_float: true
---


```{r setup, include = FALSE, echo = FALSE}
tmp = rprojroot::find_rstudio_root_file()
knitr::opts_chunk$set(root.dir = tmp, error = TRUE)
knitr::opts_knit$set(echo = TRUE, error = TRUE, root.dir = tmp)
source(paste0(tmp, "/data/environment_vars.R"))
require(raster)
require(rgdal)
require(rgeos)
require(spatstat)
# require(sf)
# require(lulcc)

require(ggplot2)
# require(psych)
rm(tmp)
```



# Data import and visualization (FF 4.3.3)
bbb

*Opuntia humifusa*

<div class = "red">
In the book, the cactus boundaries filename is *cactus_boundary.csv*, however the file provided with the book's supplemental materials is *cactus_boundaries.csv*.
</div>

<div class = "red">
The supplemental data files do not contain *cactus_boundary.csv*.
</div>

```{r prepare data, error = TRUE}
cactus = read.csv(paste0(book_data, "/cactus.csv"), header = T)

# this fails
# boundary = read.csv(paste0(book_data, "/cactus_boundary.csv"), header = T)
boundary = read.csv(paste0(book_data, "/cactus_boundaries.csv"), header = T)

ppp_window = owin(xrange = c(boundary$Xmin, boundary$Xmax), yrange = c(boundary$Ymin, boundary$Ymax))
ppp_cactus = ppp(cactus$East, cactus$North, window = ppp_window)


plot(ppp_cactus)
plot(raster::density(ppp_cactus))
summary(ppp_cactus)

# no such file
boundary_poly = readOGR(paste0(book_data, "/cactus_boundaries"))

```

## Cactus positions vs. CSR?

Use quadrat functions in spatstat:

```{r}
q_cactus = quadratcount(ppp_cactus, nx = 4, ny = 4)
plot(q_cactus)
str(q_cactus, 1)

quadrat.test(ppp_cactus, nx = 4, ny = 4)

c(q_cactus)
hist(q_cactus)
```


# Univariate Point Patterns (FF 4.3.4)

## K and L functions

No edge effect correction.


```{r}

k_cactus = Kest(ppp_cactus, correction = "none")
l_cactus = Lest(ppp_cactus, correction = "none")


plot(k_cactus)
plot(l_cactus)
# k_cactus



```


## K and L with edge correction



```{r}

k_cactus_i = Kest(ppp_cactus, correction = "isotropic")
l_cactus_i = Lest(ppp_cactus, correction = "isotropic")
k_cactus_t = Kest(ppp_cactus, correction = "translate")
l_cactus_t = Lest(ppp_cactus, correction = "translate")

{
  par(mfrow = c(3, 2))
  
  plot(k_cactus, main = "No correction")
  plot(l_cactus, main = "No correction")
  plot(k_cactus_i, main = "Isotropic correction")
  plot(l_cactus_i, main = "Isotropic correction")
  plot(k_cactus_t, main = "Translation correction")
  plot(l_cactus_t, main = "Translation correction")
}


```

## L-Functions normalized to 0

I haven't used the syntax in the second argument before but it's kind of elegant, if a bit opaque at first.


```{r}
plot(l_cactus_i, . - r ~ r)

```


## Ggplot with Kest and Lest?

```{r}
str(l_cactus_i, 1)

ggplot(l_cactus_i, aes(x = r, y = iso)) +
  geom_line() +
  geom_line(aes(x = r, y = theo), colour = "red") +
  xlab("Radius") + ylab("L")

ggplot(l_cactus_i, aes(x = r, y = iso - r)) +
  geom_line()



```


## Dispersion significance

Page 118

MCMC simulation for CSR rejection region.


Figure 4.7a includes a global envelope

```{r}

csr_env_cactus = envelope(
  ppp_cactus, 
  fun = Lest,
  nsim = 99,
  rank = 1,
  correction = "iso",
  global = FALSE)

csr_env_cactus_global = envelope(
  ppp_cactus, 
  fun = Lest,
  nsim = 99,
  rank = 1,
  correction = "isotropic",
  global = TRUE)

str(csr_env_cactus_global, 1)
# {
  # par(mfrow = c(2, 1))
  plot(csr_env_cactus, . - r ~ r, shade = c("hi", "lo"), legend = F, xlab = "normalized L", main = "L-function with MCMC envelope")
  plot(csr_env_cactus, shade = c("hi", "lo"), legend = F, xlab = "normalized L", main = "L-function with MCMC envelope")
  plot(csr_env_cactus_global, shade = c("hi", "lo"))
  # plot(csr_env_cactus_none, . - r ~ r, shade = c("hi", "lo"))
# }


```

## Pair correlation function

From page 119

Are the `translate` and `translation` options different?

From the `Kest` help entry it seems that they are...

Figure 4.7b has an additional curve with higher smoothing using a `stoyan` value higher than 0.15.

Their values are much lower near r = 0.

```{r}


pair_cactus_t = pcf(ppp_cactus, correction = "translate")
pair_cactus_t = pcf(ppp_cactus, correction = "translation")
pair_cactus_env = envelope(ppp_cactus, pcf, nsim = 99, rank = 1, correction = "translation", global = FALSE)
pair_cactus_env_2 = envelope(ppp_cactus, pcf, nsim = 99, rank = 1, correction = "translation", global = FALSE, funargs = list(stoyan = 0.4))
plot(pair_cactus_t, legend = F)
plot(pair_cactus_env, legend = F)
plot(pair_cactus_env_2, legend = F)

{
plot(pair_cactus_env, xlim = c(0.2, 14), ylim = c(0, 2.5), legend = F, shade = c("lo", "hi"), ylab = "radius", main = "pair correlation function (g)")
points(obs ~ r, data = pair_cactus_env_2, type = "l")
}


```





## Nearest neighbor: G-function


Y-axis is the probability that the nearest-neighbor is at most r units away. 
I.e. that the nearest-neighbor is closer than r units.

X-axis range is much smaller than for `K(r)` because the nearest-neighbor for any cacti are no more than 4 m away.


G(r) is higher than the 95% confidence envelope in the range of 2 to 3 meters.

High values for G(r) mean that more points have shorter nearest-neighbor distances than expected.
The probability of points being closer together than r is higher than in CSR suggesting aggregation.


Adding the ECDF doesn't add additional info and makes the plot harder to interpret (in my opinion).


```{r}
g_env_cactus = pair_cactus_env = envelope(ppp_cactus, Gest, nsim = 99, rank = 1, correction = "rs", global = FALSE)

nn_cactus = nndist(ppp_cactus)
nn_cactus

plot(g_env_cactus, main = "G-function"); plot(ecdf(nn_cactus), add = T)
```



# Marked Point Patterns - section 4.3.5



